Avoiding global deforestation by taxing land in agricultural production: The implications for global markets

### 2024-03-14 paper.Rmd
Update paper.Rmd

<!-- where $V_1$ is a vector of all the variables in both systems, i.e., all variables in GTAP and all variables in GTM. Of import, $V_1$ includes the land change variables, which are exogenous to the GTAP model. $V_1$ is then used by the GTAP model (represented by function \mbox{GTAP}) and its outputs are passed to GTM as a vector of variables $V_2$. This output vector contains the change in rents, which is treated as exogenous in the GTM model. GTM then processes the $V_2$ variables as inputs and passes its solution back to GTAP as $V_1$. To solve the system simultaneously, we require that $V_1$ and $V_2$ be equal. With several iterations, the system converges to the change in land and land rent that is consistent across the models, i.e., the land change results in the land rent changes that, when entered into the GTM model, result in the same land changes as assumed by the GTAP model. -->

# Changes under the baseline projections

To understand the potential for policy actions to achieve net zero global deforestation, an important first step was creating a realistic baseline scenario that depicts the likely changes over the decade under study, 2023--2033. Data for this were drawn from CEPII [@CEPII2024]. To build our scenario, we use two of their variables that we determined to be key: population and GDP. The CEPII projections do not contain any information on land use. To build a more complete baseline, we also include the projections on forested land change included in the GTM. Using the GTAP model in connection with the GTM model (to assure that the changes in land are consistent between the two models), we calculate the overall change in factor quantity/productivity growth required to achieve the targeted values in 2033. In Table \@ref(tab:baseline-growth), we show the projected population and GDP growth per region. GDP increases for all regions of the world with the largest percentage changes occurring in South Asia (+308.6%), China (+189.8%), and Sub-Saharan Africa (+172.9%). Population also is projected to grow in most regions with the largest changes happening in Sub-Saharan Africa (+55.7%). Japan, Russia, and East Asia are projected to see their populations decline by 10.4%, 4.7%, and 2.6%, respectively. As for the projected change in agricultural land, our baseline projections show a significant increase in Oceania (+27.6%), Rest of Latin America (+25.8%), Sub-Saharan Africa (+17.4%), Central America (+13.9%), and Brazil (+13.9%). Japan is the only nation that sees a significant decline (-34.9%) in agricultural land, which is likely tied in part to population decline.

@@

Focusing on the absolute area of the forestland, as generated by the GTM model, we estimate that between 2023 and 2033, 144.2 million hectares (3.6%) of global forests will be converted to agricultural lands with the largest decline expected to happen in Sub-Saharan Africa, which is estimated to lose 39.7 million hectares (7.0%) of its forestland (Table \@ref(tab:baseline-forest-change)). Southeast Asia, Brazil, and the rest of South America also see forestland shrink, with decreases of 22.6 million hectares (9.3%), 19.0 million hectares (3.8%), and 23.5 million hectares (6.7%), respectively. Conversely, the United States is estimated to add 1.3 million hectares (0.5%) to its forestland.

@@

Against the backdrop of the baseline scenario, we turn to the question of what policy could prevent global deforestation in the period of 2023–2033. Because the area of forestland is mostly determined by the relative returns of land in agriculture and forestry, in our policy scenario we modify the return to land in agriculture by applying a uniform global tax on the use of land in the production of agricultural output. Solving the two models together shows that it would require a tax of roughly 70 percent to be assessed on the use of land for the deforestation rate to fall globally to zero. The economic impact of such a policy measure would not be limited to the cost of the tax, it would also impact food production and its prices. For example, with this policy in place, most regions of the world would see significant declines in the amount of agricultural land in 2033 relative to the baseline scenario where land conversion was not taxed. For the United States, there would be 2.1 percent less agricultural land (Table 3). Larger decreases would happen in other major agricultural production regions such as China (-2.5 percent), South Asia (-7.9 percent), Brazil (-8.6 percent), and Russia (-9.3 percent).

@@

A visual depiction of how this policy would impact global trade is presented in Figure 1. The United States would be expected to see its agricultural exports increase significantly. Conversely, the regions where agricultural exports would be expected to drop the most are East Asia, South Asia, Southeast Asia, and Central America, which would see decreases of 14.5 percent, 5.9 percent, 5.1 percent, and 5.0 percent, respectively.

@@

Overall, the tax policy, which results in zero net deforestation globally, has a negative economic impact on all regions except Japan and Canada, which see an increase in consumer well-being of
$977.1 million and $166.9 million, respectively (Table 5). Consumer well-being is a measurement of equivalent variation and here measures the benefit/harm consumers experience in terms of income as a result of the policy changes. The United States, along with 10 other regions, see well-being drop by more than $1 billion. The most negatively impacted regions are South Asia (-$28.0 billion), China (-$14.5 billion), and Southeast Asia (-$13.6 billion). In total, this tax-driven policy is estimated to decrease global consumer well-being in 2033 by $92.1 billion relative to the scenario where no tax on agricultural land was in place.

@@

Much of this loss in well-being may be the result of higher prices. Japan is the only nation that sees any price impacts that are beneficial for consumers resulting from this policy, with the price of agricultural goods decreasing 0.3 percent (Table \@ref(tab:pm)). All other regions, for both agricultural and non-agricultural products, see prices rise between 0.3 and 5.1 percent.

>>>>

<!-- where $V_1$ is a vector of all the variables in both systems, i.e., all variables in GTAP and all variables in GTM. Of import, $V_1$ includes the land change variables, which are exogenous to the GTAP model. $V_1$ is then used by the GTAP model (represented by function \mbox{GTAP}) and its outputs are passed to GTM as a vector of variables $V_2$. This output vector contains the change in rents, which is treated as exogenous in the GTM model. GTM then processes the $V_2$ variables as inputs and passes its solution back to GTAP as $V_1$. To solve the system simultaneously, we require that $V_1$ and $V_2$ be equal. We focus on the relative returns of land in agriculture and forestry as they greatly impact land use changes. Thus, in our scenarios, we examine the impact of different tax rates on the use of land in the production of agricultural output. With several iterations, the system converges to the change in land and land rent that is consistent across the models, i.e., the land change results in the land rent changes that, when entered into the GTM model, result in the same land changes as assumed by the GTAP model. -->

# Changes under the zero tax scenario

Our first scenario examines the likely changes over the decade under study, 2023--2033. To build our scenario, we use data from CEPII [@CEPII2024] and focus on two of their variables that we determined to be key: population and GDP. In Table \@ref(tab:baseline-growth), we show the projected population and GDP growth per region. GDP increases for all regions of the world with the largest percentage changes occurring in South Asia (+308.6%), China (+189.8%), and Sub-Saharan Africa (+172.9%). Population also is projected to grow in most regions with the largest changes happening in Sub-Saharan Africa (+55.7%). Japan, Russia, and East Asia are projected to see their populations decline by 10.4%, 4.7%, and 2.6%, respectively. As the CEPII projections do not contain any information on land use, we also include the projections on forested land change included in the GTM. Using the GTAP model in connection with the GTM model (to assure that the changes in land are consistent between the two models), we then calculate the overall change in factor quantity/productivity growth required to achieve the targeted values in 2033 while assuming a uniform zero percent tax on the use of land in the production of agricultural output. Results suggest that there will be a significant increase in agricultural land in Oceania (+27.6%), Rest of Latin America (+25.8%), Sub-Saharan Africa (+17.4%), Central America (+13.9%), and Brazil (+13.9%). Japan is the only nation that sees a significant decline (-34.9%) in agricultural land, which is likely tied in part to population decline.

@@

Focusing on the absolute area of the forestland, as generated by the GTM model, we estimate that under this scenario between 2023 and 2033, 144.2 million hectares (3.6%) of global forests will be converted to agricultural lands with the largest decline expected to happen in Sub-Saharan Africa, which is estimated to lose 39.7 million hectares (7.0%) of its forestland (Table \@ref(tab:baseline-forest-change)). Southeast Asia, Brazil, and the rest of South America also see forestland shrink, with decreases of 22.6 million hectares (9.3%), 19.0 million hectares (3.8%), and 23.5 million hectares (6.7%), respectively. Conversely, the United States is estimated to add 1.3 million hectares (0.5%) to its forestland.

@@

Our second scenario examines the impact of a tax of roughly 70 percent on the use of agricultural land. Solving the two models together shows that the economic impact would not be limited to the cost of the tax, it would also impact food production and its prices. For example, most regions of the world would see significant declines in the amount of agricultural land in 2033 relative to the baseline scenario where land conversion was not taxed. For the United States, there would be 2.1 percent less agricultural land (Table 3). Larger decreases would happen in other major agricultural production regions such as China (-2.5 percent), South Asia (-7.9 percent), Brazil (-8.6 percent), and Russia (-9.3 percent). This would be because the global net deforestation rate would be brought to zero and 144.2 million hectares that would otherwise have been converted to agricultural land by 2033 would be preserved as forestland.

@@

A visual depiction of how this tax would impact global trade is presented in Figure 1. The United States would be expected to see its agricultural exports increase significantly. Conversely, the regions where agricultural exports would be expected to drop the most are East Asia, South Asia, Southeast Asia, and Central America, which would see decreases of 14.5 percent, 5.9 percent, 5.1 percent, and 5.0 percent, respectively.

@@

Overall, the 70 percent tax, which results in zero net deforestation globally, is liekly to have a negative economic impact on all regions except Japan and Canada, which see an increase in consumer well-being of
$977.1 million and $166.9 million, respectively (Table 5). Consumer well-being is a measurement of equivalent variation and here measures the benefit/harm consumers experience in terms of income as a result of the policy changes. The United States, along with 10 other regions, are estimated to see well-being drop by more than $1 billion. The most negatively impacted regions would be South Asia (-$28.0 billion), China (-$14.5 billion), and Southeast Asia (-$13.6 billion). In total, this tax-driven policy is estimated to decrease global consumer well-being in 2033 by $92.1 billion relative to the scenario where no tax on agricultural land was in place.

@@

Much of this loss in well-being may be the result of higher prices. Japan is the only nation that is estimated to see any price impacts that are beneficial for consumers, with the price of agricultural goods decreasing 0.3 percent (Table \@ref(tab:pm)). All other regions, for both agricultural and non-agricultural products, see prices rise between 0.3 and 5.1 percent.
